---
title: "cBioportal Boxplot: GE ~ Clinical (Dohoon)"
output:
  html_document:
    toc: yes
  html_notebook:
    code_folding: hide
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, fig.height = 6, fig.width = 6)
library(cgdsr)
library(survival) 
library("ggpubr") # fancy box/jitter/violin plots

## Functions
box_jitter_plot <- function(expr, cliclass, 
                            labels=  c("Positive", "Negative"), 
                            my_genes=c("SCLY", "SLC7A11", "SEPHS2", "SEPSECS", "PSTK"),
                            ylab="Expression (Z-score)"
                            ){
    # jitter plot for two groups
    # expr: df containing both clinical and expression data, row are samples, columns are genes(expression), clinical features (ER/Stage ...)
    # cliclass: colname for the clinical group of interest, so that x-axis groups of the boxplot will be stratified bases on values in this col
    # labels: select values in cliclass to be considered, sometimes there are many values, including "", NA, 'unequivacal', and use this option we can exclude them and only keep, e.g. ER "Positive", "Negative"
    
    # print information
    nb_samples <- table(expr[, cliclass])
    print(paste('Number of samples for', cliclass))
    print(nb_samples)
    
    print('t.test was used for p-value calculation')
    print(paste(c("comparison: ", labels)))
    
    # Boxplot + jitterplot (always include pvalue)
    ggboxplot(expr, x = cliclass,
              select =labels,
              order = labels,
              y = my_genes,
              combine = T,
              color = cliclass, palette = "jco",
              ylab = ylab, # can adjust
              xlab = F,
              add = "jitter", add.params = list(size = 0.1, jitter = 0.2)) +
        stat_compare_means(comparisons = list(labels), method = "t.test") +  # Welch's T-test
        theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank())  # removed xlabs
}

violin_box_plot <- function(expr, cliclass, 
                            labels=  c("Positive", "Negative"), 
                            my_genes=c("SCLY", "SLC7A11", "SEPHS2", "SEPSECS", "PSTK"),
                            ylab="Expression (Z-score)"
                            ){
    # Similar to box_jitter_plot, you can simply change ggboxplot to ggviolin
    # difference: excluded print information and function description, add = boxplot
    
    # Boxplot + jitterplot (always include pvalue)
    ggviolin(expr, x = cliclass,
              select =labels,
              order = labels,
              y = my_genes,
              combine = T,
              color = cliclass, palette = "jco",
              ylab = ylab, # can adjust
              xlab = F,
              add = "boxplot"
    ) +
        stat_compare_means(comparisons = list(labels), method = "t.test")  # Welch's T-test
}


box_jitter_plot_multiple_groups <- function(expr, cliclass, 
                                   labels=  c("Positive", "Negative"), 
                                   my_genes=c("PSTK", "SCLY", "SEPHS2", "SEPSECS", "SLC7A11"),
                                   ylab="Expression (Z-score)",
                                   xlab_angle=0, ylab_height=-2.5, 
                                   ref='.all.'){
    # box_jitter_plot for multiple_groups (n≥3)
    # expr: data.frame containing both expression and clinical data
    # cliclass: colname of clinical data to use for stratification of x-axis
    # labels: labels to be selected 
    # xlab_angle: x label rotation angle. Improves visualization, if xlab too long
    # ylab_height: the height to show gloabal p-value (one-way anova)
    # steps: 
    # 1. Levene's test (failed for Do's dataset, skipped)
    # 2. Pair-wise Welch t.test against ref
    
    # print information
    print(paste('Number of samples for', cliclass))
    print(table(expr[, cliclass]))
    print(paste(' Welch t.test was used for p-value calculation, * marks deviation from mean of ref: ', ref))
    
    # Levene's test (pre-ANOVA)
    test <- function(gene, expr, cliclass){
        leveneTest(expr[,gene] ~ expr[, cliclass], data = expr)
    }
    levene_res <- apply(data.frame(my_genes), 1, test, expr=expr, cliclass=cliclass)
    print(levene_res)

    
    
    # Boxplot + jitterplot
    ggboxplot(expr, x = cliclass,
              select =labels, order = labels,
              y = my_genes,
              combine = T,  xlab = F, ylab = ylab,
              color = cliclass, palette = "jco",
              add = "jitter", add.params = list(size = 0.1, jitter = 0.2))+
        #stat_compare_means(method = "anova", label.y = ylab_height)+        # Add global annova p-value
        stat_compare_means(label = "p.signif", method = "t.test", ref.group = ref, hide.ns = TRUE)  + # compare with ref
        rotate_x_text(angle = xlab_angle)  #+
    #theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank()) # removed xlabs
}

violin_box_plot_multiple_groups <- function(expr, cliclass, 
                                   labels=  c("Positive", "Negative"), 
                                   my_genes=c("PSTK", "SCLY", "SEPHS2", "SEPSECS", "SLC7A11"),
                                   ylab="Expression (Z-score)",
                                   xlab_angle=0, ylab_height=-2.5, ref='.all.'){
    # box_jitter_plot for multiple_groups (n≥3)
    # expr: data.frame containing both expression and clinical data
    # cliclass: colname of clinical data to use for stratification of x-axis
    # labels: labels to be selected 
    # xlab_angle: x label rotation angle. Improves visualization, if xlab too long
    # ylab_height: the height to show gloabal p-value (one-way anova)
    # steps: 
    # 1. Levene's test
    # 2. Anova or Welch t.test
    
    # print information
    print(paste('Number of samples for', cliclass))
    print(table(expr[, cliclass]))
    print('anova  was used for p-value calculation, * marks deviation from mean of all')
    
    # Levene's test (pre-ANOVA)
    # test <- function(gene, expr, cliclass){
    #     leveneTest(expr[,gene] ~ expr[, cliclass], data = expr)
    # }
    # levene_res <- apply(data.frame(my_genes), 1, test, expr=expr, cliclass=cliclass)
    # print(levene_res)

    # Boxplot + jitterplot
    ggboxplot(expr, x = cliclass,
              select =labels,
              order = labels,
              xlab = F,
              y = my_genes,
              combine = T,
              color = cliclass, palette = "jco",
              ylab = ylab,
              add = "jitter", add.params = list(size = 0.1, jitter = 0.2))+
        #stat_compare_means(method = "anova", label.y = ylab_height)+        # Add global annova p-value
        stat_compare_means(label = "p.signif", method = "t.test", ref.group = ref, hide.ns = TRUE)  + # compare with ref
        rotate_x_text(angle = xlab_angle)  +
        theme(
            axis.title.x=element_blank(), 
            axis.text.x=element_blank(), 
            axis.ticks.x=element_blank()) # removed xlabs
}
```


# Communication
- Visualize and test association between clinical data with gene expression (z-scores)
- p-values are results of wilcox.test between two groups of expression
- z-scores larger than 6 are capped to 6, for better visualization

- Database: CBioportal
- cancer: Breast Cancer
- subset: Breast Cancer Breast Invasive Carcinoma (TCGA, Cell 2015), brca_tcga_pub2015_all, 818 samples
- expression profile: brca_tcga_pub2015_rna_seq_v2_mrna_median_Zscores

All files could be downloaded from [here](./).


# Load Expression Data
```{r}
library(cgdsr)
mycgds <- CGDS("http://www.cbioportal.org/public-portal/")

studies <- getCancerStudies(mycgds)
#studies[grepl("breast invasive carcinoma", studies[, 2], ignore.case = TRUE), 1:2]  # test
study <- "brca_tcga_pub2015"
print(paste('study:', study))

cases <- getCaseLists(mycgds, study)
case <- "brca_tcga_pub2015_all"
print(paste('case:', case))


profiles <- getGeneticProfiles(mycgds, study)
profile <- c("brca_tcga_pub2015_rna_seq_v2_mrna_median_Zscores")
print(paste('profile:', profile))

my_genes <- c("SCLY", "SLC7A11", "SEPHS2", "SEPSECS", "PSTK")

ExpData <- getProfileData(mycgds, genes = my_genes, 
                       geneticProfiles = profile,
                       caseList = case)
```

## Overview of Expression Data
```{r}
# remove outlier of PSTK > 20
print('set outliers with expression > 6 to 6')
print(paste(c('num ourliers before capping:', sum(ExpData>6, na.rm = T))))
ExpData <- replace(ExpData, ExpData > 6, 6)
print(paste(c('num ourliers after capping:', sum(ExpData>6, na.rm = T))))


# overview
par(mfrow=c(3,2))

hist(ExpData$SCLY, 100 )
hist(ExpData$SLC7A11, 100)
hist(ExpData$SEPHS2, 100)
hist(ExpData$SEPSECS, 100)
hist(ExpData$PSTK, 100)
#ggdensity(ExpData, x = c("SCLY", "SLC7A11", "SEPHS2", "SEPSECS", "PSTK"), merge = T )
```


# Load Clinical Data
```{r}
clinicaldata <- getClinicalData(mycgds, caseList = case)

ClinicData <- clinicaldata[, c("AJCC_PATHOLOGIC_TUMOR_STAGE", "CANCER_TYPE_DETAILED", 
                               #"NTE_ER_STATUS", "NTE_HER2_STATUS", "NTE_PR_STATUS_BY_IHC", # 2 classes
                               "ER_STATUS_BY_IHC", "HER2_FISH_STATUS", "PR_STATUS_BY_IHC"  # 3 classes, exept HER2 (4)
                               )]

## "" to NA for ER/HER2/PR status
ClinicData[ClinicData == "Equivocal"] <- NA
ClinicData[ClinicData == ""] <- NA

## Remove 'stage' from AJCC_PATHOLOGIC_TUMOR_STAGE
ClinicData$AJCC_PATHOLOGIC_TUMOR_STAGE <- gsub("Stage ", "", ClinicData$AJCC_PATHOLOGIC_TUMOR_STAGE)

## Find TNBC (Triple Neg, vs others, excluding NA)
ClinicData$TNBC <- ifelse(
    ClinicData$ER_STATUS_BY_IHC=="Negative" 
    & ClinicData$HER2_FISH_STATUS=="Negative" 
    & ClinicData$PR_STATUS_BY_IHC=="Negative", "TNBC", "Others")
table(ClinicData$TNBC)

## check sample_ids
stopifnot(all(rownames(ExpData) %in% rownames(ClinicData)))
stopifnot(all(rownames(ClinicData) %in% rownames(ExpData)))

## sort row.names
ClinicData <- ClinicData[row.names(ExpData), ]
expr <- merge(ClinicData, ExpData, by = 0)
```


# Plotting

## TNBC
```{r fig.height = 8, fig.width = 6, fig.align = "center"}
box_jitter_plot(expr,"TNBC", c("Others", "TNBC"))
violin_box_plot(expr,"TNBC", c("Others", "TNBC"))
```

## ER_STATUS_BY_IHC
```{r fig.height = 8, fig.width = 6, fig.align = "center"}
box_jitter_plot(expr, "ER_STATUS_BY_IHC")
violin_box_plot(expr,"ER_STATUS_BY_IHC")
```

## HER2_FISH_STATUS
```{r fig.height = 8, fig.width = 6, fig.align = "center"}
box_jitter_plot(expr, "HER2_FISH_STATUS")
violin_box_plot(expr,"HER2_FISH_STATUS")
```


## PR_STATUS_BY_IHC
```{r fig.height = 8, fig.width = 6, fig.align = "center"}
box_jitter_plot(expr, "PR_STATUS_BY_IHC")
violin_box_plot(expr,"PR_STATUS_BY_IHC")
```


# ANOVA

## TUMOR_STAGE
```{r, fig.height = 8, fig.width = 15, fig.align = "center"}
print("passed Levene's test, perform ANOVA")

cliclass="AJCC_PATHOLOGIC_TUMOR_STAGE"
my_genes=c("PSTK", "SCLY", "SEPHS2", "SEPSECS", "SLC7A11")
labels = c("I", "IA",  "IB",   
           "II", "IIA", "IIB", 
           "III", "IIIA", "IIIB", "IIIC", 
           "IV", "X")
xlab_angle = 80 
ylab_height = -3
ylab = 'Expression (Z-score)'
ref='I'

print(paste('ref is ', ref))

# print information
nb_samples <- table(expr[, cliclass])
print(paste('Number of samples for', cliclass))
print(nb_samples)

print('Welch t.test was used for p-value calculation')
print(paste(c("comparison: ", labels)))

res <- compare_means(c(PSTK, SCLY, SEPHS2, SEPSECS, SLC7A11) ~ CANCER_TYPE_DETAILED,
                     data = expr, method = 't.test')
sig_comparisons <- data.frame(res[res$p.signif != 'ns',])
print(sig_comparisons)

# Boxplot + jitterplot
ggboxplot(expr, x = cliclass,
          select =labels, order = labels,
          y = my_genes,
          combine = T,  xlab = F, ylab = ylab,
          color = cliclass, 
          add = "jitter", add.params = list(size = 0.1, jitter = 0.2))+
    #stat_compare_means(method = "anova", label.y = ylab_height)+        # Add global annova p-value
    stat_compare_means(label = "p.signif", method = "t.test", ref.group = ref, hide.ns = TRUE)  + # compare with ref
    rotate_x_text(angle = xlab_angle)  #+
#theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank()) # removed xlabs

# Violin plot + boxplot
ggviolin(expr, x = cliclass,
          select =labels, order = labels,
          y = my_genes,
          combine = T,  xlab = F, ylab = ylab,
          color = cliclass, 
          add = "boxplot", add.params = list(size = 0.1, jitter = 0.2))+
    #stat_compare_means(method = "anova", label.y = ylab_height)+        # Add global annova p-value
    stat_compare_means(label = "p.signif", method = "t.test", ref.group = ref, hide.ns = TRUE)  + # compare with ref
    rotate_x_text(angle = xlab_angle)  #+
#theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank()) # removed xlabs
        
```


## CANCER_TYPE_DETAILED 

## Show all diff to '.all.'
```{r, fig.height = 12, fig.width = 12, fig.align = "center"}
box_jitter_plot_multiple_groups(
    expr, 
    cliclass="CANCER_TYPE_DETAILED", 
    my_genes=c("PSTK", "SCLY", "SEPHS2", "SEPSECS", "SLC7A11"),
    labels = c("Breast Invasive Lobular Carcinoma",
    "Invasive Breast Carcinoma",
    "Breast Invasive Ductal Carcinoma",
    "Breast Mixed Ductal and Lobular Carcinoma"),
    xlab_angle = 80, 
    ylab_height = -3, 
    ylab = 'Expression (Z-score)',
    ref='Breast Invasive Lobular Carcinoma')

# violin_box_plot_anova(expr, "CANCER_TYPE_DETAILED", 
#                       c("Breast Invasive Lobular Carcinoma",
#                         "Invasive Breast Carcinoma",
#                         "Breast Invasive Ductal Carcinoma",
#                         "Breast Mixed Ductal and Lobular Carcinoma"),
#                       xlab_angle=80,
#                       ylab_height = -3)
```
```{r}

```


<!-- ## Show all sig pair-wise comparisons -->
<!-- ```{r, fig.height = 12, fig.width = 12, fig.align = "center"} -->
<!-- box_jitter_plot_multiple_groups <- function(expr, cliclass,  -->
<!--                                    labels=  c("Positive", "Negative"),  -->
<!--                                    my_genes=c("PSTK", "SCLY", "SEPHS2", "SEPSECS", "SLC7A11"), -->
<!--                                    ylab='Expression (log2 intensity)', -->
<!--                                    xlab_angle=0, ylab_height=-2.5, ref='.all.'){ -->
<!--     # box_jitter_plot for multiple_groups (n≥3) -->
<!--     # expr: data.frame containing both expression and clinical data -->
<!--     # cliclass: colname of clinical data to use for stratification of x-axis -->
<!--     # labels: labels to be selected  -->
<!--     # xlab_angle: x label rotation angle. Improves visualization, if xlab too long -->
<!--     # ylab_height: the height to show gloabal p-value (one-way anova) -->
<!--     # steps:  -->
<!--     # 1. Levene's test -->
<!--     # 2. Anova or Welch t.test -->

<!--     # print information -->
<!--     print(paste('Number of samples for', cliclass)) -->
<!--     print(table(expr[, cliclass])) -->
<!--     print('anova  was used for p-value calculation, * marks deviation from mean of all') -->

<!--     # Levene's test (pre-ANOVA) -->
<!--     # test <- function(gene, expr, cliclass){ -->
<!--     #     leveneTest(expr[,gene] ~ expr[, cliclass], data = expr) -->
<!--     # } -->
<!--     # levene_res <- apply(data.frame(my_genes), 1, test, expr=expr, cliclass=cliclass) -->
<!--     # print(levene_res) -->

<!--     # pair-wise t.test (Welch) -->
<!--     #res <- compare_means(c(PSTK, SCLY, SEPHS2, SEPSECS, SLC7A11) ~ CANCER_TYPE_DETAILED,  -->
<!--     #                     data = expr, method = 't.test') -->
<!--     #sig_comparisons <- data.frame(res[res$p.signif != 'ns',][, c('group1', 'group2')]) -->
<!--     #sig_comparisons <- split(sig_comparisons,seq(nrow(sig_comparisons))) -->
<!--     #sig_comparisons <- as.vector(t(sig_comparisons)) -->

<!--     # Boxplot + jitterplot -->
<!--     ggboxplot(expr, x = cliclass, -->
<!--               select =labels, -->
<!--               order = labels, -->
<!--               xlab = F, -->
<!--               y = my_genes, -->
<!--               combine = T, -->
<!--               color = cliclass, palette = "jco", -->
<!--               ylab = ylab, -->
<!--               add = "jitter", add.params = list(size = 0.1, jitter = 0.2))+ -->
<!--         #stat_compare_means(method = "anova", label.y = ylab_height)+        # Add global annova p-value -->
<!--         #stat_compare_means(label = "p.signif", method = "t.test", ref.group = ref, hide.ns = TRUE)  + -->
<!--         rotate_x_text(angle = xlab_angle)  + -->
<!--         theme( -->
<!--             axis.title.x=element_blank(),  -->
<!--             axis.text.x=element_blank(),  -->
<!--             axis.ticks.x=element_blank()) + # removed xlabs -->
<!--         stat_compare_means(comparisons =  list(labels[c(1,2)], labels[c(1,3)],labels[c(1,4)], -->
<!--                                                labels[c(2,3)],labels[c(2,4)], -->
<!--                                                labels[c(3,4)]),  -->
<!--                            label = "p.signif", hide.ns = T,  -->
<!--                            method = "t.test" )  # Welch's T-test -->
<!-- } -->

<!-- box_jitter_plot_multiple_groups(expr, cliclass="CANCER_TYPE_DETAILED",  -->
<!--                                 my_genes=c("PSTK", "SCLY", "SEPHS2", "SEPSECS", "SLC7A11"), -->
<!--                       labels = c("Breast Invasive Lobular Carcinoma", -->
<!--                         "Invasive Breast Carcinoma", -->
<!--                         "Breast Invasive Ductal Carcinoma", -->
<!--                         "Breast Mixed Ductal and Lobular Carcinoma"), -->
<!--                       ref='Breast Invasive Lobular Carcinoma') -->

<!-- # violin_box_plot_anova(expr, "CANCER_TYPE_DETAILED",  -->
<!-- #                       c("Breast Invasive Lobular Carcinoma", -->
<!-- #                         "Invasive Breast Carcinoma", -->
<!-- #                         "Breast Invasive Ductal Carcinoma", -->
<!-- #                         "Breast Mixed Ductal and Lobular Carcinoma"), -->
<!-- #                       xlab_angle=80, -->
<!-- #                       ylab_height = -3) -->
<!-- ``` -->
