---
title: "cBioportal Survival curve(Dohoon)"
output:
  html_document:
    toc: yes
  html_notebook:
    code_folding: hide
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, fig.height = 6, fig.width = 6)
library(cgdsr)
library(survival)

density_scatter_plot <- function(x, y, xname="x", yname="y", max=2, method="pearson"){
    # https://stackoverflow.com/questions/17093935/r-scatter-plot-symbol-color-represents-number-of-overlapping-points 
    
    # FOR CORRELATION
    x1 <- x
    x2 <- y
    res <- cor.test(x1, x2, method = method, use = 'complete.obs')
    p <- res$p.value
    correlation <- res$estimate
    
    # FOR plotting
    ## Data in a data.frame
    x1[x1 > max] <- max
    x2[x2 > max] <- max
    df <- data.frame(x1,x2)

    ## Use densCols() output to get density at each point
    x <- densCols(x1,x2, colramp=colorRampPalette(c("black", "white")))
    df$dens <- col2rgb(x)[1,] + 1L
    
    ## Map densities to colors
    cols <-  colorRampPalette(c("#000099", "#00FEFF", "#45FE4F", 
                                "#FCFF00", "#FF9400", "#FF3100"))(256)
    df$col <- cols[df$dens]
    
    ## Plot it, reordering rows so that densest points are plotted on top
    plot(x2~x1, data=df[order(df$dens),], 
         pch=20, col=col, cex=2, 
         main="Scatterplot", 
         xlab=xname, ylab=yname
         
         )
    legend(x='bottomright', legend=paste('Corr =', round(correlation, 2), 
                                         "\npvalue =", formatC(p, format = 'e', 1)))
}

survival_plot <- function(ClinicData, ExpData, GeneNames, main=" "){
    # ClinicData: Survival Data
    # ExpData: expression data
    # GeneNames: Gene of interest, e.g. c("gene1", "gene2"), c("gene1")
    
    # Prep
    rank_cut <- floor(dim(ExpData)[1] * 0.25)
    
    if (length(GeneNames) == 1){
        GeneName <- GeneNames[1]
        
        bottom <- rank(ExpData[, GeneName]) < rank_cut
        bottom <- row.names(ExpData[bottom, ] )
        
        top <- rank(-ExpData[, GeneName]) < rank_cut
        top <- row.names(ExpData[top, ] )
        
        ClinicData[, "UpDown"] = NaN
        ClinicData[top, "UpDown"] = TRUE
        ClinicData[bottom, "UpDown"] = FALSE # top True, bottom False, Others NA
    }else if (length(GeneNames >= 2)) {
        GeneName <- paste(GeneNames, collapse = '.')  # for legend
        
        # list of genes
        top <- apply(-ExpData[, GeneNames], 2, rank) < rank_cut
        top <- apply(top, 1, prod) # top in all genes in GeneNames
        top <- row.names(ExpData[top==1, ])
        
        bottom <- apply(ExpData[, GeneNames], 2, rank) < rank_cut
        bottom <- apply(bottom, 1, prod)
        bottom <- row.names(ExpData[bottom==1, ])
        
        ClinicData[, "UpDown"] = NaN
        ClinicData[top, "UpDown"] = TRUE
        ClinicData[bottom, "UpDown"] = FALSE # top True, bottom False, Others NA 
    }else {
        stop("GeneNames input error")
    }
      
    # Survival  
    ## test
    sdf <- survdiff(Surv(OS_MONTHS,VITAL_STATUS1)~UpDown, data=ClinicData, rho=1) ## rho=1, Gehan-Wilcoxon test
    pvalue <- 1 - pchisq(sdf$chisq, length(sdf$n) - 1)
    
    ## plot
    fit <- survfit(Surv(OS_MONTHS,VITAL_STATUS1)~UpDown, data=ClinicData)
    plot(fit, mark.time=TRUE, col=c('blue', 'red'), lty = c(1, 2, 0), 
         xlab="Months Survival", 
         ylab="Surviving", 
         main=paste(main, "Overall Survival Kaplan-Meier Estimate", sep = "\n"))
    legend("topright", col=c('blue', 'red'), lty = c(1,2,0),
           legend = c(paste(GeneName, "Down"), 
                      paste(GeneName, "Up"), 
                      paste('pvalue =', formatC(pvalue, format = 'e', 1))
                      )
           )
    
    # Disease Free Survival
    ## test
    sdf <- survdiff(Surv(DFS_MONTHS,DFS_STATUS1)~UpDown, data=ClinicData, rho=1) ## rho=1, Gehan-Wilcoxon test
    pvalue <- 1 - pchisq(sdf$chisq, length(sdf$n) - 1)
    
    ## plot
    fit <- survfit(Surv(DFS_MONTHS,DFS_STATUS1)~UpDown, data=ClinicData)
    plot(fit, mark.time=TRUE, col=c('blue', 'red'), lty = c(1, 2, 0), 
         xlab="Months Disease Free", 
         ylab="Disease Free", 
         main= paste(main, "Disease Free Survival Kaplan-Meier Estimate", sep = "\n"))
    legend("topright", col=c('blue', 'red'), lty = c(1,2,0),
           legend = c(paste(GeneName, "Down"), 
                      paste(GeneName, "Up"), 
                      paste('pvalue =', formatC(pvalue, format = 'e', 1))
                      )
           )
}
```

# communication
- Test Overall Survival for Gene Upregulated Patients
- Up: with higher level of mRNA of such gene in the patient's canver sample compared to normal samples (Z-score top 25%)
- Down: (Z-score bottom 25%)

- Survival curves

- Database: CBioportal
- cancer: Breast Cancer
- subset: Breast Cancer Breast Invasive Carcinoma (TCGA, Cell 2015), brca_tcga_pub2015_all, 818 samples
- expression profile: brca_tcga_pub2015_rna_seq_v2_mrna_median_Zscores

# Note

All files could be downloaded from [here](./).

# Load Expression Data
```{r}
library(cgdsr)
mycgds <- CGDS("http://www.cbioportal.org/public-portal/")
studies <- getCancerStudies(mycgds)
studies[grepl("breast invasive carcinoma", studies[, 2], ignore.case = TRUE), 1:2]  # test
study <- "brca_tcga_pub2015"

cases <- getCaseLists(mycgds, study)
case <- "brca_tcga_pub2015_all"

profiles <- getGeneticProfiles(mycgds, study)
profile <- c("brca_tcga_pub2015_rna_seq_v2_mrna_median_Zscores")

data <- getProfileData(mycgds, genes = c("SCLY", "SLC7A11", 
                                         "SEPHS2", "SEPSECS", "PSTK"), 
                       geneticProfiles = profile,
                       caseList = case)
```

## Overview of Expression Data
```{r}
par(mfrow=c(3,2))
hist(data$SCLY, 100 )
hist(data$SLC7A11, 100, xlim = range(-2,10))
hist(data$SEPHS2, 100)
hist(data$SEPSECS, 100)
hist(data$PSTK, 100)

clinicaldata <- getClinicalData(mycgds, caseList = case)
```

# Load Survival Data
```{r}
library(survival)
## The OS MONTHS is the overall Survival duration of patient after the first surgery
## The OS STATUS is the event at OS MONTHS. 
## The survival plot works only with studies that have non empty OS MONTHS and OS STATUS.
head(clinicaldata[, c("OS_MONTHS", "VITAL_STATUS")])

ClinicData <- clinicaldata[, c("OS_MONTHS", "VITAL_STATUS")]
ClinicData <- clinicaldata[, c("DAYS_TO_DEATH","DFS_MONTHS", "DFS_STATUS", "VITAL_STATUS", "OS_MONTHS", "OS_STATUS")]
ClinicData$VITAL_STATUS1 <- ClinicData$VITAL_STATUS ==  "Dead"  # Dead -> True
ClinicData$DFS_STATUS1 <- ClinicData$DFS_STATUS ==  "Recurred/Progressed"  # Recurred -> True

paste("There are", sum(ClinicData$VITAL_STATUS1), "patients Dead" )
paste("There are", sum(ClinicData$DFS_STATUS1), "patients recurred")

stopifnot(all(rownames(data) %in% rownames(ClinicData)))
stopifnot(all(rownames(ClinicData) %in% rownames(data)))
```


# Filter by expression (Up vs Down)
- Up: ranked top 25%
- Down: ranked bottom 25%

## TOX genes

### SCLY Upregulation
```{r}
survival_plot(ClinicData = ClinicData, ExpData = data, GeneNames = c('SCLY'), main = "Breast Invasive Carcinoma (TCGA)")
```

### SLC7A11 Upregulation
```{r}
survival_plot(ClinicData = ClinicData, ExpData = data, GeneNames = c('SLC7A11'), main = "Breast Invasive Carcinoma (TCGA)")
```
### SLC7A11 and SCLY Upregulation 
```{r}
survival_plot(ClinicData = ClinicData, ExpData = data, GeneNames = c('SCLY', 'SLC7A11'), main = "Breast Invasive Carcinoma (TCGA)")
```

## DETOX genes
### SEPHS2 Upregulation
```{r}
survival_plot(ClinicData = ClinicData, ExpData = data, GeneNames = c('SEPHS2'), main = "Breast Invasive Carcinoma (TCGA)")
```
### SEPSECS Upregulation
```{r}
survival_plot(ClinicData = ClinicData, ExpData = data, GeneNames = c('SEPSECS'), main = "Breast Invasive Carcinoma (TCGA)")
```
### PSTK Upregulation
```{r}
survival_plot(ClinicData = ClinicData, ExpData = data, GeneNames = c('PSTK'), main = "Breast Invasive Carcinoma (TCGA)")
```

### SEPSECS.and.SEPHS2
```{r}
survival_plot(ClinicData = ClinicData, ExpData = data, GeneNames = c('SEPSECS', 'SEPHS2'), main = "Breast Invasive Carcinoma (TCGA)")
```
### SEPSECS.and.PSTK
```{r}
survival_plot(ClinicData = ClinicData, ExpData = data, GeneNames = c('SEPSECS', 'PSTK'), main = "Breast Invasive Carcinoma (TCGA)")
```
### SEPHS2.and.PSTK
```{r}
survival_plot(ClinicData = ClinicData, ExpData = data, GeneNames = c('SEPHS2', 'PSTK'), main = "Breast Invasive Carcinoma (TCGA)")
```
### SEPHS2.and.SEPSECS.and.PSTK
```{r}
survival_plot(ClinicData = ClinicData, ExpData = data, GeneNames = c('SEPHS2','SEPSECS', 'PSTK'), main = "Breast Invasive Carcinoma (TCGA)")
```

## Others
### SLC7A11.and.SEPHS2 Survival
```{r}
survival_plot(ClinicData = ClinicData, ExpData = data, GeneNames = c('SLC7A11', 'SEPHS2'), main = "Breast Invasive Carcinoma (TCGA)")


```

### SLC7A11 and SEPHS2 Correlation
```{r}
density_scatter_plot(x = data$SLC7A11, y = data$SEPHS2,  max=2, 
                     xname = "SLC7A11", yname = "SEPHS2", 
                     method = 'pearson')
```





