---
title: "DEseq2"
author: "RuiLi"
date: "07/24/2019"
output:
  html_document:
    toc: yes
    code_folding: hide
    toc_float: yes
---

# Communication
- For Merav
- ATACseq

# Notes:
- Annotated to promoter regions only (+/- 5000bp of TSS), overlap of gene_body is not annotated


# Code LOG:
- included annotation
- included merging
- used TPM normalization
- save seperate results into excel rather than csv to prevent gene name bug in excel
- print cut-offs
- export results in original order, not ordered by LFC nor FDR (more robust)
- export significant results, too (FDR and |LFC|, not including edges)
- better VolcanoPlot by setting Max Y to 50 (>50 set to 50)
- better html output, skip nb, use html
- both OneBigModel and SeperatedModels
- fixed row.names (gencode id) bug
- output both FPKM and TPM, 06/28/2019
- both shrinked LFC and raw LFC in same output Excel, and both plots
- add lib-size
- added COUNT excel output
- Pvalue -> FDR in Volcano-plot, ylim 0,50, xlim fixing, smaller dots


```{r setup, include=FALSE}
library(DESeq2)
library(ggplot2)
library("scater")
library(WriteXLS)
require(plyr)
library(gdata)

thresh_p <- 0.05
thresh_LFC <- 1

volcanoplot <- function (res, 
                         lfcthresh=2, sigthresh=thresh_p, 
                         name='name', 
                         legendpos="topright", labelsig=FALSE, textcx=1, ...) {
    res_ <- res
    res_$padj[res_$padj < 1e-10 & !is.na(res_$padj)] <- 1e-10 # y-axis top value 50
    
    main=paste("Volcano Plot", name, sep='\n')
    with(res_, plot(log2FoldChange, -log10(padj), pch=20, main=main,  ylim=c(0,10), ...))
    with(subset(res_, padj<sigthresh ), 
         points(log2FoldChange, -log10(padj), pch=20, col="orange",...))
    with(subset(res_, abs(log2FoldChange)>lfcthresh), 
         points(log2FoldChange, -log10(padj), pch=20, col="green", ...))
    with(subset(res_, padj<sigthresh & abs(log2FoldChange)>lfcthresh), 
         points(log2FoldChange, -log10(padj), pch=20, col="red",...))
    if (labelsig) {
        require(calibrate)
        with(subset(res_, padj<sigthresh & abs(log2FoldChange)>lfcthresh), 
             textxy(log2FoldChange, -log10(padj), labs=Gene, cex=textcx, ...))
    }
    legend(legendpos, xjust=1, yjust=0.5, 
           legend=c(paste("FDR<",sigthresh,sep=""), 
                    paste("|LogFC|>",lfcthresh,sep=""), 
                    "both"), pch=20,
           col=c("orange","green","red"))
}


maplot <- function (res, thresh=thresh_p, labelsig=FALSE, textcx=1, ...) {
    with(res, 
         plot(baseMean, log2FoldChange, pch=20, cex=.5, log="x", ...))
    with(subset(res, padj<thresh_p), 
         points(baseMean, log2FoldChange, col="red", pch=20, cex=.5))
    if (labelsig) {
        require(calibrate)
        with(subset(res, padj<thresh_p), 
             textxy(baseMean, log2FoldChange, labs=Gene, cex=textcx, col=2))
    }
}


process_deseq_res <- function(res="lfcshrink.res", res2="results.res", name='name', anno='anno.df', norm_exp="tpm.df"){
    ## Summary
    print(name)
    print("Summary using FDR cut-off only (LFC not used)")
    summary(res, alpha=thresh_p)
    
    print("Summary using both FDR and LFC_shrinked cut-off")
    sig_idx <- res$padj<thresh_p & abs(res$log2FoldChange) > thresh_LFC
    sig_idx[is.na(sig_idx)] <- FALSE
    res_sig <- res[sig_idx,]
    print(table(sig_idx))
    
    hist(res$pvalue, breaks=50, col="grey", 
         main = paste('Histogram of p-values (un-adjusted)', name, sep = "\n"), 
         xlab = 'pvalues', ylab = 'Frequency')
    
    hist(res$padj, breaks=50, col="grey", 
         main = paste('Histogram of Q-values (adjusted)', name, sep = "\n"), 
         xlab = 'padj', ylab = 'Frequency')
    
    maplot(res, main=paste("MAplot", paste(name, "LFC_shrinked"), sep="\n"))
    maplot(res2, main=paste("MAplot", paste(name, "LFC_raw"), sep="\n"))
    
    volcanoplot(res,lfcthresh=thresh_LFC, sigthresh=thresh_p, 
                textcx=.8, name= paste(name, "LFC_shrinked"))
    volcanoplot(res2, lfcthresh=thresh_LFC, sigthresh=thresh_p, 
                textcx=.8, name= paste(name, "LFC_raw"))
    
    res.df <- as.data.frame(res)
    names(res.df)[2] <- "log2FoldChange_shrinked"
    names(res.df)[3] <- "lfcSE_shrinked"

    res2.df <- as.data.frame(res2)
    names(res2.df)[2] <- "log2FoldChange_raw"
    names(res2.df)[3] <- "lfcSE_raw"
    res2.df <- res2.df[, c(2,3)]

    resdata <- merge(res.df, res2.df, by=0, sort=F, all.x=T)
    resdata <- merge(anno, resdata, by.x=1, by.y=1, sort=FALSE, all.y=T)
    resdata <- merge(resdata, norm_exp, by.x=1, by.y=0, all.x=T)
    
    head(resdata)

    ## Write results
    # write.csv(resdata, row.names = FALSE, 
    #           file=paste(name, 'deseq.csv', sep = '.'))  # for later analysis
    WriteXLS(x = resdata, 
         ExcelFileName = paste(name, 'deseq2.xlsx', sep = '.'), 
         row.names = F, SheetNames = 'sheet1', na = '-')  # for user
    
    WriteXLS(x = resdata[sig_idx,], 
     ExcelFileName = paste(name, 'deseq2.sig.FDR', thresh_p, 
                           'LFC', thresh_LFC, 'xlsx', sep = '.'), 
     row.names = F, SheetNames = 'sheet1', na = '-')  # for user
    



}
```

# Print Cut-offs
```{r}
paste("FDR cut-off:", thresh_p)
paste("Log2FC cut-off:", thresh_LFC)
```


# Read Data
```{r}
df <- read.table('../../../snakemake/2_analysis/featureCount/merged_peak.count.txt', 
                 sep="\t", header=TRUE, 
                 row.names = 1) # row.name in cts(matrix)
colnames(df)
dim(df)
```

## Clean Up Data
```{r}
# clean names
colnames(df) <- gsub("\\.bam$", "", colnames(df))
colnames(df) <- gsub("...markDup.", "", colnames(df))
# colname_formatted <- c(
#     "KO_D0_3", "WT_D0_3", "KO_D2_3", "WT_D2_3", "KO_D8_3", "WT_D8_3",
#     "KO_D0_1", "KO_D0_2", "KO_D2_1", "KO_D2_2", "KO_D8_1", "KO_D8_2", 
#     "WT_D0_1", "WT_D0_2", "WT_D2_1", "WT_D2_2", "WT_D8_1", "WT_D8_2")
# paste(colnames(df), colname_formatted, sep = "->")
# colnames(df) <- colname_formatted

colnames(df)
head(row.names(df))
dim(df)
```

## Create cts
```{r}
cts <- df[, 6:ncol(df)]
cts <- as.matrix(cts)
colnames(cts)
head(cts)
paste("lib-size:")
colSums(cts)
```

## Filter Based on Expression
```{r}
expression_filter <- rowSums(cts) >= 1  # default 10
cts <- cts[expression_filter, ]
df <- df[expression_filter, ]
dim(cts)
dim(df)
```

# Read Anno
```{r}
anno <- read.table(
    "../../ChIPPeakAnno/merged_peaks.full_anno.csv", sep=",",
    header=T)
colnames(anno)
anno <- anno[, c("peak_id", "seqnames", "start", "end", "width",  "geneBiotype", "geneID", "geneName" )]
dim(anno)
head(anno)
```

# COUNT output
```{r}
count <- cts
count_out <- merge(anno, count,  by.x=1, by.y=0, all.y=T, sort=F)
head(count_out)
WriteXLS(x = count_out, 
         ExcelFileName = 'COUNT.xlsx', row.names = F, SheetNames = 'sheet1', na = '-')
```


# TPM calculation
```{r}
tpm <- calculateTPM(cts, df$Length)
tpm <- data.frame(tpm)
colnames(tpm) <- paste("TPM", colnames(tpm), sep = ":")
tpm_out <- merge(anno, tpm, by.x=1, by.y=0, all.y=T, sort=F)
head(tpm_out)
WriteXLS(x = tpm_out, 
         ExcelFileName = 'TPM.xlsx', row.names = F, SheetNames = 'sheet1', na = '-')
```


# FPKM calculation (Not FPKM)
```{r}
fpkm <- calculateFPKM(cts, df$Length)
fpkm <- data.frame(fpkm)
colnames(fpkm) <- paste("FPKM", colnames(fpkm), sep = ":")
fpkm_out <- merge(anno, fpkm, by.x=1, by.y=0, all.y=T, sort=F)
head(fpkm_out)
WriteXLS(x = fpkm_out, 
         ExcelFileName = 'FPKM.xlsx', row.names = F, SheetNames = 'sheet1', na = '-')
```



# DESeq Experiment Design (One Big Model)
- for QC and comparison with seperated model only
```{r design}
sample <- factor(c("KO", "KO", "Lin", "Lin", "S0_CD71_low", "S0_CD71_low", "S0_CD71_medium", "S0_CD71_medium", "S1", "S1", "S2", "S2", "S3_large", "S3_large", "S3_small", "S3_small", "S4_S5"))
sample
# 
batch <- factor(c("1", "2", "1", "2", "1", "2", "1", "2", "1", "2",
    "1", "2", "1","2", "1", "2", "2"))
batch

type <- factor(paste(sample, batch, sep = "_"))
type
#mouse <- factor(rep(c("Mouse1", "Mouse2", "Mouse3"), 6))
coldata <- data.frame(row.names=colnames(cts), 
                      sample, batch, type)
coldata
```

## Model fitting
```{r}
dds <- DESeqDataSetFromMatrix(countData = cts, 
                              colData = coldata, 
                              design = ~  sample
                              )  # converted to alph-order
# dds$type <- relevel(dds$type, 
#                     ref = "DMSO_2D"
#                     )
# dds$mouse <- relevel(dds$mouse, 
#                      ref = "Mouse1"
#                      )

dds
dds <-DESeq(dds)
resultsNames(dds)
saveRDS(dds, file = 'dds.oneBigModel.rds')
```


# QC Plots

## Data transformation
```{r}
vsd <- vst(dds, blind=FALSE)

#rld <- rlog(dds, blind=FALSE)
counts <- counts(dds, normalized=0)
logCounts <- log10(counts +1 )

normed <- counts(dds, normalized=1)
logNormed <- log10(normed+1)
```

## Histogram of normalized data
```{r}
hist(logNormed, main='log10(x+1) normalized expression')
```

## Dispersion plot
```{r}
plotDispEsts(dds, main="Dispersion plot")
```


## PCA plots
```{r}
pcaData <- plotPCA(vsd, intgroup=c("sample", "batch"), returnData=TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))

ggplot(pcaData, aes(PC1, PC2, color=sample, shape=batch)) +
  geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed()
```


### With label
```{r}
library(ggplot2)
library(ggrepel)

pcaplot <- ggplot(pcaData, aes(PC1, PC2, color=sample, shape=batch)) +
    geom_point(size=3) +
    #geom_text(aes(label=type),hjust=1, vjust=0) +
    xlab(paste0("PC1: ",percentVar[1],"% variance")) +
    ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
    coord_fixed() +
    coord_fixed(ratio = percentVar[1]/percentVar[2]/4, 
                xlim = NULL, ylim = NULL, expand = T,
                clip = "off")

### geom_label_repel
pcaplot + 
  geom_label_repel(aes(label = sample),
                  box.padding   = 0.35, 
                  point.padding = 0.5,
                  segment.color = 'grey50') +
  theme_classic()
```



## Sample Heatmap
```{r}
library("RColorBrewer")
library('pheatmap')
library("PoiClaClu")
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
poisd <- PoissonDistance(t(counts(dds)))
samplePoisDistMatrix <- as.matrix( poisd$dd ) 
rownames(samplePoisDistMatrix) <- paste( coldata$type, sep="-" ) 
colnames(samplePoisDistMatrix) <- NULL 
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(samplePoisDistMatrix,
         clustering_distance_rows=poisd$dd,
         clustering_distance_cols=poisd$dd,
         col=colors, 
         clustering_method='complete'
        )

```

# Results from different contrasts

## One Big Model (S0_CD71_medium vs S0_CD71_low)
```{r}
#resultsNames(dds)
name <- "DM-S0_CD71_medium_vs_S0_CD71_low"
contrast <- c('sample', 'S0_CD71_medium', 'S0_CD71_low')
res <- lfcShrink(dds, contrast = contrast, type = 'ashr')
res2 <- results(dds, contrast = contrast)
process_deseq_res(res = res, res2=res2, name = name, anno = anno, norm_exp = tpm)
```

## One Big Model (S1 vs S0_CD71_medium)
```{r}
#resultsNames(dds)
name <- "S1_vs_S0_CD71_medium"
contrast <- c('sample', 'S1', 'S0_CD71_medium')
res <- lfcShrink(dds, contrast = contrast, type = 'ashr')
res2 <- results(dds, contrast = contrast)
process_deseq_res(res = res, res2=res2, name = name, anno = anno, norm_exp = tpm)
```

## One Big Model (S2 vs S1)
```{r}
#resultsNames(dds)
name <- "S2_vs_S1"
contrast <- c('sample', 'S2', 'S1')
res <- lfcShrink(dds, contrast = contrast, type = 'ashr')
res2 <- results(dds, contrast = contrast)
process_deseq_res(res = res, res2=res2, name = name, anno = anno, norm_exp = tpm)
```

## One Big Model (S3_large vs S2)
```{r}
#resultsNames(dds)
name <- "S3_large_vs_S2"
contrast <- c('sample', 'S3_large', 'S2')
res <- lfcShrink(dds, contrast = contrast, type = 'ashr')
res2 <- results(dds, contrast = contrast)
process_deseq_res(res = res, res2=res2, name = name, anno = anno, norm_exp = tpm)
```

## One Big Model (S4_S5 vs S3_small)
```{r}
#resultsNames(dds)
name <- "S4_S5_vs_S3_small"
contrast <- c('sample', 'S4_S5', 'S3_small')
res <- lfcShrink(dds, contrast = contrast, type = 'ashr')
res2 <- results(dds, contrast = contrast)
process_deseq_res(res = res, res2=res2, name = name, anno = anno, norm_exp = tpm)
```

## One Big Model (S0_CD71_low vs KO)
```{r}
#resultsNames(dds)
name <- "S0_CD71_low_vs_KO"
contrast <- c('sample', 'S0_CD71_low', 'KO')
res <- lfcShrink(dds, contrast = contrast, type = 'ashr')
res2 <- results(dds, contrast = contrast)
process_deseq_res(res = res, res2=res2, name = name, anno = anno, norm_exp = tpm)
```

## One Big Model (S0_CD71_medium vs KO)
```{r}
#resultsNames(dds)
name <- "S0_CD71_medium_vs_KO"
contrast <- c('sample', 'S0_CD71_medium', 'KO')
res <- lfcShrink(dds, contrast = contrast, type = 'ashr')
res2 <- results(dds, contrast = contrast)
process_deseq_res(res = res, res2=res2, name = name, anno = anno, norm_exp = tpm)
```

## One Big Model (S1 vs KO)
```{r}
#resultsNames(dds)
name <- "S1_vs_KO"
contrast <- c('sample', 'S1', 'KO')
res <- lfcShrink(dds, contrast = contrast, type = 'ashr')
res2 <- results(dds, contrast = contrast)
process_deseq_res(res = res, res2=res2, name = name, anno = anno, norm_exp = tpm)
```

<!-- ## Seperated Model (DM-G_vs_DM4) less DEG, discarded -->
<!-- ```{r} -->
<!-- #resultsNames(dds) -->
<!-- #idx <- grep('D7', coldata$type) -->
<!-- idx <- c(1, 2, 3,  7, 8, 9) -->
<!-- head(cts[, idx]) -->
<!-- coldata[idx,] -->
<!-- dds <- DESeqDataSetFromMatrix(countData = cts[, idx],  -->
<!--                               colData = coldata[idx,],  -->
<!--                               design =  ~ type)  # converted to alph-order -->
<!-- dds -->
<!-- dds <-DESeq(dds) -->
<!-- dds -->
<!-- resultsNames(dds) -->

<!-- name <- "DM-G_vs_DM4" -->
<!-- contrast <- c('type', 'DM_G', 'DM4') -->
<!-- res <- lfcShrink(dds, contrast = contrast, type = 'ashr') -->
<!-- res2 <- results(dds, contrast = contrast) -->
<!-- process_deseq_res(res = res, res2=res2, name = name, anno = anno, norm_exp = tpm) -->
<!-- ``` -->
