---
title: "DEseq2"
author: "RuiLi"
date: "05/16/2018"
output:
  html_document:
    toc: yes
    code_folding: hide
    toc_float: yes
---

# Communication


# Code LOG:
- included annotation
- included merging
- used TPM normalization
- save seperate results into excel rather than csv to prevent gene name bug in excel
- print cut-offs
- export results in original order, not ordered by LFC nor FDR (more robust)
- export significant results, too (FDR and |LFC|, not including edges)
- better VolcanoPlot by setting Max Y to 50 (>50 set to 50)
- better html output, skip nb, use html
- both OneBigModel and SeperatedModels

```{r setup, include=FALSE}
library(DESeq2)
library(ggplot2)
library("scater")
library(WriteXLS)
require(plyr)
library(gdata)

thresh_p <- 0.05
thresh_LFC <- 1

volcanoplot <- function (res, 
                         lfcthresh=2, sigthresh=thresh_p, 
                         name='name', 
                         legendpos="topright", labelsig=FALSE, textcx=1, ...) {
    res_ <- res
    res_$pvalue[res_$pvalue < 1e-50 & !is.na(res_$pvalue)] <- 1e-50 # y-axis top value 50
    
    main=paste("Volcano Plot", name, sep='\n')
    with(res_, plot(log2FoldChange, -log10(pvalue), pch=20, main=main,  ylim=c(0,50), ...))
    with(subset(res_, padj<sigthresh ), 
         points(log2FoldChange, -log10(pvalue), pch=20, col="orange", ...))
    with(subset(res_, abs(log2FoldChange)>lfcthresh), 
         points(log2FoldChange, -log10(pvalue), pch=20, col="green", ...))
    with(subset(res_, padj<sigthresh & abs(log2FoldChange)>lfcthresh), 
         points(log2FoldChange, -log10(pvalue), pch=20, col="red", ...))
    if (labelsig) {
        require(calibrate)
        with(subset(res_, padj<sigthresh & abs(log2FoldChange)>lfcthresh), 
             textxy(log2FoldChange, -log10(pvalue), labs=Gene, cex=textcx, ...))
    }
    legend(legendpos, xjust=1, yjust=0.5, 
           legend=c(paste("FDR<",sigthresh,sep=""), 
                    paste("|LogFC|>",lfcthresh,sep=""), 
                    "both"), pch=20,
           col=c("orange","green","red"))
}

maplot <- function (res, thresh=thresh_p, labelsig=FALSE, textcx=1, ...) {
    with(res, 
         plot(baseMean, log2FoldChange, pch=20, cex=.5, log="x", ...))
    with(subset(res, padj<thresh_p), 
         points(baseMean, log2FoldChange, col="red", pch=20, cex=1.5))
    if (labelsig) {
        require(calibrate)
        with(subset(res, padj<thresh_p), 
             textxy(baseMean, log2FoldChange, labs=Gene, cex=textcx, col=2))
    }
}


process_deseq_res <- function(res, name='name', anno='anno(df)'){
    ## Summary
    print(name)
    print("Summary using FDR cut-off only (LFC not used)")
    summary(res, alpha=thresh_p)
    
    print("Summary using both FDR and LFC cut-off")
    sig_idx <- res$padj<thresh_p & abs(res$log2FoldChange) > thresh_LFC
    sig_idx[is.na(sig_idx)] <- FALSE
    res_sig <- res[sig_idx,]
    print(table(sig_idx))
    
    ## Output
    # ## Order by LFC (up to down)
    # res <- res[order(res$log2FoldChange, decreasing = T), ]
    ## Merge with normalized count data
    resdata <- merge(as.data.frame(res), anno, 
                     by.x=0, by.y=1, sort=FALSE, all.x=T)
    names(resdata)[1] <- "Gene"
    head(resdata)
    ## Write results
    # write.csv(resdata, row.names = FALSE, 
    #           file=paste(name, 'deseq.csv', sep = '.'))  # for later analysis
    WriteXLS(x = resdata, 
         ExcelFileName = paste(name, 'deseq2.xlsx', sep = '.'), 
         row.names = F, SheetNames = 'sheet1', na = '-')  # for user
    
    WriteXLS(x = resdata[sig_idx,], 
     ExcelFileName = paste(name, 'deseq2.sig.FDR', thresh_p, 
                           'LFC', thresh_LFC, 'xlsx', sep = '.'), 
     row.names = F, SheetNames = 'sheet1', na = '-')  # for user
    
    ## Hist of P-values/FDR
    ## Examine plot of p-values
    hist(res$pvalue, breaks=50, col="grey", 
         main = paste('Histogram of p-values (un-adjusted)', name, sep = "\n"), 
         xlab = 'pvalues', ylab = 'Frequency')
    
    hist(res$padj, breaks=50, col="grey", 
         main = paste('Histogram of p-values (adjusted)', name, sep = "\n"), 
         xlab = 'padj', ylab = 'Frequency')
    
    maplot(resdata, main=paste("MAplot", name, sep="\n"))
    volcanoplot(resdata, lfcthresh=thresh_LFC, sigthresh=thresh_p, 
                textcx=.8, xlim=c(-2.3, 2), name=name)

}
```

# Print Cut-offs
```{r}
paste("FDR cut-off:", thresh_p)
paste("Log2FC cut-off:", thresh_LFC)
```


# Read Data
```{r readdata}
df <- read.table('../feature_count/counts.gene_id.s0.txt', 
                 sep="\t", header=TRUE, 
                 row.names = 1) # row.name in cts(matrix)
colnames(df)
dim(df)
```


## Clean Up Data
```{r}
# clean names
colnames(df) <- gsub("\\.bam$", "", colnames(df))
colnames(df) <- gsub("sorted_reads.", "", colnames(df))
# colname_formatted <- c(
#     "KO_D0_3", "WT_D0_3", "KO_D2_3", "WT_D2_3", "KO_D8_3", "WT_D8_3",
#     "KO_D0_1", "KO_D0_2", "KO_D2_1", "KO_D2_2", "KO_D8_1", "KO_D8_2", 
#     "WT_D0_1", "WT_D0_2", "WT_D2_1", "WT_D2_2", "WT_D8_1", "WT_D8_2")
# paste(colnames(df), colname_formatted, sep = "->")
# colnames(df) <- colname_formatted

row.names(df) <- gsub("\\.*", "", row.names(df))  # for annotation purpose
colnames(df)
dim(df)
```

## Create cts
```{r}
cts <- df[, 6:ncol(df)]
cts <- as.matrix(cts)
colnames(cts)
head(cts)
```

## Filter Based on Expression
```{r}
expression_filter <- rowSums(cts) >= 1  # default 10
cts <- cts[expression_filter, ]
df <- df[expression_filter, ]
dim(cts)
dim(df)
```

# Read Anno
```{r}
anno <- read.table(
    "~/github/bpipes/rnaseq/annotations/gencode.hu.v29/gencode.v29.primary_assembly.anno.txt.gz", 
    header=T)
anno$Gene <- gsub("\\.*", "", anno$Gene)  # for annotation purpose

dim(anno)
head(anno)
```


# TPM calculation
```{r}
tpm <- calculateTPM(cts, df$Length)
tpm <- data.frame(tpm)
colnames(tpm) <- paste("TPM", colnames(tpm), sep = ":")
head(tpm)
tpm_out <- merge(anno, tpm, by.x=1, by.y=0, all.y=T)
head(tpm_out)
WriteXLS(x = tpm_out, 
         ExcelFileName = 'TPM.xlsx', row.names = F, SheetNames = 'sheet1', na = '-')
```



# DESeq Experiment Design (One Big Model)
- for QC and comparison with seperated model only
```{r design}
genotype <- factor(c(
    rep("CR",6), 
    rep("KD", 6)
    ))
genotype

time <- factor(c(rep(c("D7", "D14"), each=3), 
               rep(c("D7", "D14"), each=3)))
time

type <- factor(paste(genotype, time, sep = "_"))
type
#mouse <- factor(rep(c("Mouse1", "Mouse2", "Mouse3"), 6))
coldata <- data.frame(row.names=colnames(cts), 
                      type, genotype, time)
coldata
```

## Model fitting
```{r}
dds <- DESeqDataSetFromMatrix(countData = cts, 
                              colData = coldata, 
                              design = ~ type
                              )  # converted to alph-order
# dds$type <- relevel(dds$type, 
#                     ref = "DMSO_2D"
#                     )
# dds$mouse <- relevel(dds$mouse, 
#                      ref = "Mouse1"
#                      )

dds
dds <-DESeq(dds)
resultsNames(dds)
saveRDS(dds, file = 'dds.oneBigModel.rds')
```


# QC Plots

## Data transformation
```{r}
vsd <- vst(dds, blind=FALSE)

#rld <- rlog(dds, blind=FALSE)
counts <- counts(dds, normalized=0)
logCounts <- log10(counts +1 )

normed <- counts(dds, normalized=1)
logNormed <- log10(normed+1)
```

## Histogram of normalized data
```{r}
hist(logNormed, main='log10(x+1) normalized expression')
```

<!-- ## Dispersion plot -->
<!-- ```{r} -->
<!-- plotDispEsts(dds, main="Dispersion plot") -->
<!-- ``` -->


## PCA plots
- look at batch effect 
```{r}
pcaData <- plotPCA(vsd, intgroup=c("genotype", "time"), returnData=TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))

ggplot(pcaData, aes(PC1, PC2, color=genotype, shape=time)) +
  geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed()
```

## Heatmap
- samples clustered by time, then by batch
```{r}
library("RColorBrewer")
library('pheatmap')
library("PoiClaClu")
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
poisd <- PoissonDistance(t(counts(dds)))
samplePoisDistMatrix <- as.matrix( poisd$dd ) 
rownames(samplePoisDistMatrix) <- paste( coldata$type, sep="-" ) 
colnames(samplePoisDistMatrix) <- NULL 
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(samplePoisDistMatrix,
         clustering_distance_rows=poisd$dd,
         clustering_distance_cols=poisd$dd,
         col=colors, 
         clustering_method='complete'
        )

```

# Results from different contrasts

## One Big Model (D7_KD_vs_CR_v1)
```{r}
resultsNames(dds)
res <- lfcShrink(dds, contrast = c('type', 'KD_D7', 'CR_D7'), type = 'ashr')
name <- "D7_KD_vs_CR_v1"
process_deseq_res(res = res, name = name, anno = tpm_out)
```

## One Big Model (D14_KD_vs_CR_v1)
```{r}
resultsNames(dds)
res <- lfcShrink(dds, contrast = c('type', 'KD_D14', 'CR_D14'), type = 'ashr')
name <- "D14_KD_vs_CR_v1"
process_deseq_res(res = res, name = name, anno = tpm_out)
```



## D7_KD_vs_CR (Seperated Models)
```{r}
idx <- grep('D7', coldata$type)
# idx <- c(1, 2, 5, 6)
head(cts[, idx])
coldata[idx,]
dds <- DESeqDataSetFromMatrix(countData = cts[, idx], 
                              colData = coldata[idx,], 
                              design = ~ type )  # converted to alph-order
dds <-DESeq(dds)
dds
resultsNames(dds)

res <- lfcShrink(dds, contrast = c('type', 'KD_D7', 'CR_D7'), type = 'ashr')
name <- "D7_KD_vs_CR"
process_deseq_res(res = res, name = name, anno = tpm_out[c(1:3, 3+idx)])
```


## D14_KD_vs_CR (Seperated Models)
```{r}
idx <- grep('D14', coldata$type)
# idx <- c(1, 2, 5, 6)
head(cts[, idx])
coldata[idx,]
dds <- DESeqDataSetFromMatrix(countData = cts[, idx], 
                              colData = coldata[idx,], 
                              design = ~ type )  # converted to alph-order
dds <-DESeq(dds)
dds
resultsNames(dds)

res <- lfcShrink(dds, contrast = c('type', 'KD_D14', 'CR_D14'), type = 'ashr')
name <- "D14_KD_vs_CR"
process_deseq_res(res = res, name = name, anno = tpm_out[c(1:3, 3+idx)])
```