---
title: "Heatmap for DEG: complex heatmap"
author: "RuiLi"
date: "01/23/2019"
output: html_document
fig_width: 3
fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width=12, fig.height=8) 

```

# Notes
- Color indicates log2(x+1) transformed normalized expression
- only significant DEGs with |LFC|>1 and FDR < 0.01 included

# Read Data
```{r,  results="hide"}
outname <- 'Heatmap of DEG'
fname <- 'deseq.csv'

df <- read.table(fname, header = TRUE, na.strings = 'NA', sep = ',')

# filter and format
thr_p <- 0.01
thr_lfc <- 1

df$padj[is.na(df$padj)] <- 1  # remove NA
sum(is.na(df$hat2vsWT6padj))
sum(is.na(df$hat2vsWT6log2FoldChange))

small_p <- df$padj < thr_p
large_fc <- abs(df$log2FoldChange) > thr_lfc

sum(small_p)
sum(large_fc)

sig <- (small_p & large_fc)
sum(sig)

df <- df[sig, ]
colnames(df)

colnames(df)[10:13] = c('4SA-TAZ.1', "4SA-TAZ.2", "Vector.1", "Vector.2")
```

## Filter Data
```{r}
# select expression columns
df_ <- df[, c(10:13)]
colnames(df_)

values_sig <- df_
df_sig <- df

logNormed_sig <- log10(values_sig + 1)  # log transformation

num_sig <- dim(df_sig)[1]
paste(num_sig, "significant genes")
```


## Heatmap for DEG 
```{r, echo=FALSE, results="hide"}
# source("https://bioconductor.org/biocLite.R")
# biocLite("ComplexHeatmap")
library(ComplexHeatmap)
library(circlize)


mat <- logNormed_sig
row.names(mat) <- row.names(df_sig)
max_exp <- max(logNormed_sig)

labels <- as.character(df_sig$symbol)  # with duplicates, bug?

markers <- c('TRPC6')
marker_idx <- match(tolower(markers), tolower(labels))

labels[marker_idx]

header = data.frame(type = c('4SA-TAZ', "4SA-TAZ", "Vector", "Vector"))

ha = HeatmapAnnotation(df = header,  
                       col = list(type = c("4SA-TAZ" =  "red", 
                                           'Vector' = 'blue'
                                           )))
```


## Heatmap Sort by LFC (Log2 Fold Change)
```{r , fig.height = 6, fig.width = 4, fofig.align = "center", echo=FALSE, results="hide"}
Heatmap(mat, 
        column_title = outname,
        row_title = paste(num_sig, 'DEG'),
        name = 'Exp',
        
        cluster_rows = F,
        cluster_columns = F, 
        clustering_method_rows = 'centroid', 
        
        show_row_names = F,
        show_column_names = T,
        
        show_row_dend = T, 
        show_column_dend = T,
        
        top_annotation = ha, 
        
        col = colorRamp2(c(0, max_exp/2, max_exp), c("blue", 'white',  "red")), 
        fontsize(4)
        ) + 
  rowAnnotation(link = row_anno_link(at = marker_idx, labels = markers),
                width = unit(5, "mm") + 
                  max_text_width(markers))

```

## Heatmap Hierarchical Clustering by Complete Distance
```{r, fig.height = 6, fig.width = 4, fofig.align = "center", echo=FALSE, results="hide"}
Heatmap(mat, 
        column_title = outname,
        row_title = paste(num_sig, 'DEG'),
        name = 'Exp',
        
        cluster_rows = T,
        cluster_columns = F, 
        clustering_method_rows = 'complete', 
        
        show_row_names = F,
        show_column_names = T,
        
        show_row_dend = T, 
        show_column_dend = T,
        
        top_annotation = ha, 
        
        col = colorRamp2(c(0, max_exp/2, max_exp), c("blue", 'white',  "red")), 
        fontsize(4)
        ) + 
  rowAnnotation(link = row_anno_link(at = marker_idx, labels = markers),
                width = unit(5, "mm") + 
                  max_text_width(markers))

```


## Heatmap Hierarchical Clustering by median Distance
```{r, fig.height = 6, fig.width = 4, fofig.align = "center", echo=FALSE, results="hide"}
Heatmap(mat, 
        column_title = outname,
        row_title = paste(num_sig, 'DEG'),
        name = 'Exp',
        
        cluster_rows = T,
        cluster_columns = F, 
        clustering_method_rows = 'median', 
        
        show_row_names = F,
        show_column_names = T,
        
        show_row_dend = T, 
        show_column_dend = T,
        
        top_annotation = ha, 
        
        col = colorRamp2(c(0, max_exp/2, max_exp), c("blue", 'white',  "red")), 
        fontsize(4)
        ) + 
  rowAnnotation(link = row_anno_link(at = marker_idx, labels = markers),
                width = unit(5, "mm") + 
                  max_text_width(markers))

```





