---
title: "Heatmap for DEG: complex heatmap"
author: "RuiLi"
date: "06/18/2019"
output:
  html_document:
    toc: yes  # table of content
    code_folding: hide
    toc_float: yes
fig_width: 3
fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width=12, fig.height=8) 
library(ComplexHeatmap)
library(circlize)
```

## Communications
- For Mandy
- Visualized with Zscore
- Only significant Differentially Expressed Genes (DEG) with FDR < 0.05 included

## Code log:
- Use zscore for each gene scaled among samples for visualization

## Read Data
```{r}
outname <- 'Heatmap of DEG'

df <- readxl::read_xlsx( '../DESeq2_batch/High_vs_Low.deseq2.sig.FDR.0.05.LFC.0.xlsx', na='-')  # na term important
df <- data.frame(df)  #important

# sort by LFC
df <- df[order(df$log2FoldChange), ]
exp_idx <- grepl("TPM", colnames(df))
colnames(df) <- gsub("TPM.", "", colnames(df))

head(df)
```


## Significance Filter (skipped)
```{r}
# sig_idx <- read.table("../1a_DE_UnionSet/DE_Union.txt") 
df.sig <- df
row.names(df.sig) <- df.sig$Name
# head(df.sig)
```

## Select expression columns, transform into Z-score
```{r}
exp_df.sig <- df.sig[, exp_idx]
colnames(exp_df.sig)
row.names(exp_df.sig)[1:5]
dim(exp_df.sig)

z.exp_df.sig <- t(scale(t(exp_df.sig)))

# num_df <- log10(exp_df + 1)  # log transformation
num_df <- z.exp_df.sig

num_sig <- dim(num_df)[1]
paste( dim(num_df)[1], "significant genes")

head(num_df)
write.csv(num_df, "sig.zscore.csv", row.names = T)
```

## Heatmap for DEG 
```{r, echo=FALSE}
max_exp <- max(num_df)
max_exp <- 2  # params
min_exp <- -2  # params
hist(num_df, 50,
     main = "Histogram of Z-score\n red-line: max in heatmaps\nblue: min")
abline(v=max_exp, col = 'red')
abline(v=min_exp, col = 'blue')


labels <- as.character(exp_df.sig$Name)  # with duplicates, bug?

markers <- c('GeneSymbol', 'GeneSymbol2')  # params
marker_idx <- match(tolower(markers), tolower(labels))

labels[marker_idx]

header = data.frame(type = rep(c('High', "Low"), each=3))

ha = HeatmapAnnotation(df = header,  
                       col = list(type = c("High" =  "orange", 
                                           'Low' = 'green'
                                           )))
```




## Heatmap Complete Hierarchical Clustering by Manhattan Distance
- Used Z-score for clustering and visualization
- Z-score calculated from TPM for each gene, scaled among samples
```{r, fig.height = 8, fig.width = 6, fofig.align = "center"}
Heatmap(num_df, 
        column_title = outname,
        row_title = paste(num_sig, 'DEG'),
        name = 'Z-score',
        
        cluster_rows = T,
        cluster_columns = F, 
        clustering_distance_rows = "manhattan",
        clustering_method_rows = 'complete', 
        
        show_row_names = F,
        show_column_names = T,
        
        show_row_dend = T, 
        show_column_dend = T,
        
        top_annotation = ha, 
        
        col = colorRamp2(c(min_exp, 0, max_exp), c("blue", 'white',  "red")), 
        fontsize(4)
        ) + 
  rowAnnotation(link = row_anno_link(at = marker_idx, labels = markers),
                width = unit(5, "mm") + 
                  max_text_width(markers))

```



<!-- ## Heatmap Sort by LFC (Log2 Fold Change) -->
<!-- ```{r , fig.height = 6, fig.width = 5, fofig.align = "center", echo=FALSE, results="hide"} -->
<!-- Heatmap(mat,  -->
<!--         column_title = outname, -->
<!--         row_title = paste(num_sig, 'DEG'), -->
<!--         name = 'Exp', -->

<!--         cluster_rows = F, -->
<!--         cluster_columns = F,  -->
<!--         clustering_method_rows = 'centroid',  -->

<!--         show_row_names = F, -->
<!--         show_column_names = T, -->

<!--         show_row_dend = T,  -->
<!--         show_column_dend = T, -->

<!--         top_annotation = ha,  -->

<!--         col = colorRamp2(c(0, max_exp/2, max_exp), c("blue", 'white',  "red")),  -->
<!--         fontsize(4) -->
<!--         ) +  -->
<!--   rowAnnotation(link = row_anno_link(at = marker_idx, labels = markers), -->
<!--                 width = unit(5, "mm") +  -->
<!--                   max_text_width(markers)) -->

<!-- ``` -->
